---
title: "Arbeidskrav 2"
output: html_document
author: Sondre Lauvdal
date: "2026-04-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown


Assignment 2

Task 1:
Load the datasets and explore the first few rows to understand the structure and types of data available. Identify and handle the missing values in both datasets. Next, merge the cleaned countries and energy data on the country column and show the first few rows of the merged dataset (Hint: Module 10, na.omit(), merge(), make sure that the column country is present in both datasets)

```{r}

# same packages as in first assignment are all loaded. Data loaded in same manner as well, then handeled with recomended functions. 

library(tidyverse) 
library(readr)
library(dplyr)
library(readxl)
library(scales)
library(ggplot2)

energy <- read_csv(
  "owid-energy-data.csv",
  show_col_types = FALSE
)

energy |>
  mutate(across(14, ~as.numeric(gsub(';','.',.)))) # to encounter problems known from loading data in assignment 1

head(energy) # head displays first part

countries <- read_xlsx('countries.xlsx')

head(countries)

countries <- countries |>  rename(country = Country, Continent = `Continent`) # changing names so it´s equal in sets

energy_cleaned   <- na.omit(energy) #na.omit "removes" missing data - it seems slightly radical in remowing rows with missing values

countries_cleaned <- na.omit(countries)

merged_data <- merge(energy_cleaned, countries_cleaned, by = "country") #datasets are merged with  country as "common factor"

head(merged_data) #shows the first few rows


```
Task 2
Calculate the average gas consumption (gas_consumption) for each continent using the merged dataset from the previous question, and then visualize this data in a plot. Discuss the insights you can draw from the visualization regarding gas consumption patterns across continents.

```{r}

#using merged data average_gas_consumption in created by continens. 
average_gas_consumption <- merged_data |> 
  group_by(Continent) |> 
  summarise(average_gas_consumption = mean(gas_consumption, na.rm = TRUE)) # average is created with summarise from dplyr reducing values per continent to one/continent


head(average_gas_consumption) #overwiew

#Creating a bar chart named "Average Gas Consumption by Continent" for visualization of average_gas_consumption
ggplot(average_gas_consumption, aes(x = Continent, y = average_gas_consumption)) + 
  geom_col(fill = "blue") + 
  labs(                     
    title = "Average Gas Consumption by Continent",
    x = "Continent",
    y = "Average Gas Consumption"
  ) +
  theme_classic()


```
Discussion task 2
Regarding graph in task 2 this doesn´t look promising... Using the merged data Africa, Asia and Oceania are not present. Seems the gas_consumption is not registered as a value per county for these continents, but at continent as a whole. 
As the graph stands - North america has a higher consumtion per country. Australia f.ex is not included in the countries file

Task 3: 
Generate descriptive statistics for gdp and total_energy_consumption for the year 2013. Then, create a scatter plot to analyze the relationship between GDP and total energy consumption in 2013. Use the total_energy_consumption variable calculated in Assignment 1 using the code below:

```{r}

#Using given code from assignment 1 in the assignment text:

consumption_vars <- names(select(merged_data, ends_with("consumption")))
consumption_vars <- consumption_vars[!(consumption_vars %in%
                                         c("fossil_fuel_consumption",
                                           "low_carbon_consumption",
                                         "renewables_consumption",
                                           "primary_energy_consumption"))]

merged_data <- merged_data %>% 
  mutate(
    total_energy_consumption = rowSums(across(all_of(consumption_vars)), na.rm = TRUE)
 )

merged_data

# Retrieving 2013 data using filter function
data_2013 <- merged_data |>  filter(year == 2013)

# Descriptive statistics. Summary generates overview of statistics
summary(data_2013[, c("gdp", "total_energy_consumption")]) 

#Creating scatter plot on the new data_2013 with GDP vs total_energy_consumption
ggplot(data_2013, aes(x = gdp, y = total_energy_consumption)) +
  geom_point(alpha = 0.8, color = "green") +
  labs(title = "GDP vs Total Energy Consumption (2013)",
       x = "GDP", y = "Total Energy Consumption") +
  theme_classic()


```
Discussion task 3
Clear trend that consumption increases with wealth.

Task 4:
Filter countries with gdp higher than the median, compute the average greenhouse_gas_emissions for these countries, and create a bar plot (Hint: Module 11, median, subset()).

```{r}

# Using subset function in order to keep rows with higher than median gdp - value from merged_data
high_gdp <- subset(merged_data, gdp > median(gdp, na.rm = TRUE)) 

#calculating average_greenhouse_gas_emissions by country in the high_gdp data made in previous line
average_greenhouse_gas_emissions <- high_gdp |> 
  group_by(country) |> 
  summarise(average_greenhouse_gas_emissions = mean(greenhouse_gas_emissions, na.rm = TRUE)) |> # as previous task summarise is used to get one value
  arrange(desc(average_greenhouse_gas_emissions)) # this one is not necessary - gives better visualisation / exploration

#creating a bar plot illustrating  average_greenhouse_gas_emissions by country
ggplot(average_greenhouse_gas_emissions, aes(x = country, y = average_greenhouse_gas_emissions)) +
  geom_bar(stat = "identity", fill = "blue") +
  labs(title = "Avg Greenhouse Gas Emissions, above average GDP countries",
       x = "Country", y = "Avg Greenshouse Gas Emissions") +
  theme_classic()


```
Task 5
Construct a histogram of gdp values using 10 bins. Choose a color that enhances visibility. Similarly, generate a histogram for gas_consumption using 10 bins. Select a contrasting color to facilitate clear visual differentiation. Next, create a scatter plot to visualize the relationship betweengdp and gas_consumption. Each point should represent a country, and use a color scale to distinguish countries. Include appropriate axes labels and a legend if necessary. Compare the insights gained from the histograms with those from the scatter plot. Discuss how the distribution characteristics align or contrast with the relationship observed in the scatter plot.

```{r}

# Starting with10 bins histogram for gdp, still working on merged data. Each bin will show how many countries fall into each gdp group - this is not for one spesific year, bot several for each country. 
ggplot(merged_data, aes(x = gdp)) +
  geom_histogram(bins = 10, fill = "blue", color = "green") +
  labs(title = "GDP", x = "GDP") + theme_classic()

#Gas Consumption histogram, basically doing the same as for gdp, also ranging over years for all countries
ggplot(merged_data, aes(x = gas_consumption)) +
  geom_histogram(bins = 10, fill = "darkorange", color = "green") +
  labs(title = "Gas Consumption", x = "Gas Consumption (TWh)") + theme_classic()

# Scatterplot with relationship between gdp and gas_consumption (colored by country)
ggplot(merged_data, aes(x = gdp, y = gas_consumption, color = country)) +
  geom_point(alpha = 0.8) +
  labs(title = "Relationship between gdp and gas_consumption", x = "GDP", y = "Gas Consumption") +
  theme_classic()

```

Discussion task 5. 
For GDP there´s no obvious distribution, though slightly more in centre, and several at 0/NA. For gas consumption it´s more bifocal. The scatterplot shows a trend with increasing consumption with increasing gdp, Brazil and France look like exeptions. 

Task 6
Fit a linear regression model predicting total_energy_consumption from gdp for countries and summarize the model. Next, plot the residuals of the linear regression model and show the r-squared value to assess its fit.


```{r}

#From chapter 23 in the textbook R has a lm function for generating linear models, also shown in lecture on models - using this one here to create linear_model from merged data predicting total_energy_consumption from gdp for countries. Looking at scatterplot for 2013 in the  relationship should be linear, keeping all years in this task. 

linear_model <- lm(total_energy_consumption ~ gdp, data = merged_data) #integrated lm model

summary(linear_model) #function for summary

# Residual plot, for the rest of the task I had to communicate with Chat gtp to make the code, functions seem standard. fitted extract estimated response on values in original dataset. residuals show difference between estimate and original value.  
plot(fitted(linear_model), residuals(linear_model),  
     xlab = "Fitted Values", ylab = "Residuals",
     main = "Residuals of linear regression model")
abline(h = 0, col = "red") #straight red line/reference line at zero

summary(linear_model)[["r.squared"]] #built in function extracts value for whole model as one number, high value indicate good model

```

continue go trough here!!!

Task 7
Create a figure with two subplots. In the first subplot display a time series of greenhouse_gas_emissions from 2000 to 2020 for Mexico, Germany, and Spain. In the second subplot display a corresponding time series of fossil_fuel_consumption for the same countries and period. Use facet_wrap() to organize the subplots by country. Analyze the plots to compare how trends in fossil fuel consumption relate to trends in greenhouse gas emissions.

```{r}

#generating a vector with relevant countries making it easier to work with them down the line, Chat gtp for hints on this one
subset_countries <- c("Mexico", "Germany", "Spain") 

#Retrieving/filtering data for three relevant countries over relevant years, naming it time series
timeseries_data <- merged_data |> 
  filter(country %in% subset_countries, year >= 2000, year <= 2020) #%in% is a logical operator to check for inclusion, found through google,


# greenhouse_gas_emissions
ggplot(timeseries_data, aes(x = year, y = greenhouse_gas_emissions, color = country)) +
  geom_line() + # line plot
  facet_wrap(~country) + # facet wrap splits in side by side subplots by variable country in this case
  labs(title = "Greenhouse Gas Emissions 2000–2020") + theme_classic()

# fossil_fuel_consumption 
ggplot(timeseries_data, aes(x = year, y = fossil_fuel_consumption, color = country)) +
  geom_line() +
  facet_wrap(~country) +
  labs(title = "Fossil Fuel Consumption 2000–2020") + theme_classic()

```

Discussion task 7
Germany is rather flat in greenhouse gas emission, but actually falls slightly in fossil consumtion - chemical or agriculture? Germany has a higher consumption than the two other countries in any case. For Mexico and Spain the curves follow for greenhouse and fossile consumption. 

Task 8
Using the merged dataset, calculate the proportion of total_energy_consumption that comes from gas_production for each country. After calculating these proportions, sort the results in descending order. Then, identify and report which country has the highest proportion of its energy consumption coming from gas, and which has the lowest

```{r}

#Creating new data gas_proportion. Then in merged data a new column with our new variable should be added through mutate. Then as in several previous tasks it´s grouped by country, then summarise is used on average to generate single value, arrange desc is asked for in the task. 
gas_proportion <- merged_data |> 
  mutate(gas_proportion = gas_production / total_energy_consumption) |> 
  group_by(country) |> 
  summarise(avg_gas_prop = mean(gas_proportion, na.rm = TRUE)) |> 
  arrange(desc(avg_gas_prop))

head(gas_proportion) #Genereates 6x2 tibble

head(gas_proportion, 1)   # Highest, picking name from 1st column in tibble equals Argentina
tail(gas_proportion, 1)   # Lowest equals Spain
```

Task 9.
How do electricity_generation, time, and Continent interact? Aggregate and visualize the total electricity generation for each continent over time. Create a line plot with different colors for each continent, and facet by continent to clearly show the trends and differences in electricity generation across continents (Hint: Module 10).


```{r}

#Still working on merged data. Grouping by time (year) and Continent before making summation where summarise is used to reduces to one value (as in average) for total electricity generation
electricity_continent <- merged_data |> 
  group_by(Continent, year) |> 
  summarise(total_electricity = sum(electricity_generation, na.rm = TRUE))

continents <- electricity_continent |> 
  distinct(Continent) |> 
  pull(Continent) # had to use chat gpt here to pick continents as distinc values, distinct(), and then pull to create a vector for iterations

#Creating a "for" iteration, as shown in module 10, in order to go through all continents ( cont ) in continents, where continents are retrieved as unique/distinct values in the line above.

for (cont in continents) {
  
  #createing data for continents where electricity continent is the summation above. It´s then filtered by Continent equel (== in iterations) to the running variable cont for iteration. 
  cont_data <- electricity_continent |> 
    filter(Continent == cont)
  
  print(
    ggplot(cont_data, aes(x = year, y = total_electricity)) +
      geom_line(color = "blue") +
      labs(title = paste("Electricity Generation —", cont),  #again I had to use chat gpt in order to figure how to plot different plots with correct names. 
           x = "year",
           y = "total electricity generation") +
      theme_classic()
  )
}

# using iterations as shown in module 10 generate different graph, and I can¨t figure out how to join them through subplots with facet().
#As far as I can understand facet anf for-loops will not work together, at least not with facet inside the loop where continets are already splitted. 
#The facet function may replace the iterations, and following is suggested as a visualization - results are comparable through the approaches. 

ggplot(electricity_continent,
       aes(x = year, y = total_electricity, color = Continent)) +
  geom_line() +
  facet_wrap(~Continent) +
  theme_classic()

```
Task 10
Conduct a regression analysis to explore the relationship between the fossil_fuel_consumption of countries and their greenhouse_gas_emissions. Discuss the results of the regression, focusing on the coefficient of the fossil_fuel_consumption variable. Interpret the significance level and the fit of the model (model_1) to assess how fossil_fuel_consumption values predict greenhouse_gas_emissions (Hint: Module 12).

```{r}

#Using lm function from earlier task and textbook for a linear model, named model_1. Datas extracted through summary

model_1 <- lm(greenhouse_gas_emissions ~ fossil_fuel_consumption, data = merged_data)

summary(model_1)

```
Discussion 10:

Lines intercept at -27,4, not to meaningful. Gas emission increase by 0.089 per unit in fossil_fuel_consumption increase. Std.error for our line/estimate is really small (0.0037), and indicate a good fit. A good fit is also reflected in a low t-value for our estimate- indication that there´s a strong relationship between gas emission and fuel consumtion, very low risk that we hit this by chance here.  

R squared is a good measure of model fit, can be read as percentage/fraction in variation/increase/decrease of one variable explained by the other. R squared here being 0,8545 means that 86% of greenhouse gas emission is explained by fuel consumtion, remaining 14 percent not explained by our model. 

Task 11
Create another model model_2 to see how coal_production impacts greenhouse_gas_emissions. Does the coal production provides a better explanation for CO2 emissions compared to the model_1 with only fossil-fuel consumption?

```{r}
#Regression in same manner as previous task. 

model_2 <- lm(greenhouse_gas_emissions ~ coal_production, data = merged_data)

summary(model_2)
```

Discussion 11
R squared is down to 0,75 in model_2 compared to 0,85 in model_1 indicating that fuel consumption explains more of variations in greenhouse gas emmisions than coal production does. Looking at residual standard error 33,91 vs 44,53 model_2 has a poorer fit.
Coal doesn´t provide a better explanation of emission. 

Task 12
Calculate and visualize the residuals for both models (model_1 and model_2). Additionally, assess the fit of each model by comparing their R-squared values. Which model better fits the data based on residual patterns and R-squared values?

```{r}
# Calculating residuals through residuals().

residual_1 <- residuals(model_1)

summary(residual_1)

residual_2 <- residuals(model_2)

summary(residual_2)

# Support from chat gpt in understanding this one. Residuals are returned as a vector, and must be converted to datafrane. Next values are fitted, compared to "x" as a common reference point i graphs, creating one dataframe/residual. Then setting up residuals as scatterplot in a new graph 

residual_df1 <- data.frame(
  fitted = fitted.values(model_1),
  residuals = residual_1
)

residual_df2 <- data.frame(
  fitted = fitted.values(model_2),
  residuals = residual_2
)

ggplot() +
  geom_point(data = residual_df1,
             aes(x = fitted, y = residuals, color = "Model 1")) +
  geom_point(data = residual_df2,
             aes(x = fitted, y = residuals, color = "Model 2")) +
  labs(
    title = "Residuals Model 1 vs Model 2",
    x = "Fitted Values",
    y = "Residuals",
    color = "Model"
  ) +
  theme_classic()

```
Discussion 12. 
R squared was discussed in task 11, and same applies here. From the scatterplot there´s a larger spread away from zero in model_2 indication a poorer fit compared to model_1

Task 13
Outline, with words, next steps you would take to continue analyzing the greenhouse_gas_emissions using the available datasets.

- Correct for number of inhabitants to asses high-consumers?
- Check for effect of eg renewables_consumption og greenhouse_gas_emission, same for nuclear?
- Production/consumption of energy for different countries?
- Assososiation between wealth and renewable use?
